Efficient pedigree recording for fast population genetics simulation

### 2018-01-15 forwards_paper.tex
merged disco edits; spelling

The idea of storing genealogical information to speed up simulations is not new.
It was implemented in AnA-FiTS~\citep{aberer2013rapid},
but without the critical step of discarding irrelevant genealogical information.
\citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
by keeping track of genealogies over the preceeding 8 generations
and only tracking neutral genotypes for those segments having descendants across this window.
Our approach is similar, but uses genealogies across the entire duration of the simulation.
The embellished pedigree is equivalent to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which has been the subject of substantial study
% under the assumptions of coalescent theory
\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, it is unclear if an ARG-based approach would share
the computational advantages of the data structures we use here~\citep{kelleher2016efficient}.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simpified in \msprime{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
compatible with exising methods for downstream analyses.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
thread. Our implementation allows a maxmimum of four simplification intervals to be in the queue at once. This
parallelized version also performed fitness calculation in parallel using two threads of execution in C++.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
and XXX.

>>>>

The idea of storing genealogical information to speed up simulations is not new.
It was implemented in AnA-FiTS~\citep{aberer2013rapid},
but without the critical step of discarding irrelevant genealogical information.
\citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
by keeping track of genealogies over the preceding 8 generations
and only tracking neutral genotypes for those segments having descendants across this window.
Our approach is similar, but uses genealogies across the entire duration of the simulation.
The embellished pedigree is equivalent to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which has been the subject of substantial study
% under the assumptions of coalescent theory
\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, it is unclear if an ARG-based approach would share
the computational advantages of the data structures we use here~\citep{kelleher2016efficient}.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simplified in \msprime{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
compatible with existing methods for downstream analyses.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
thread. Our implementation allows a maximum of four simplification intervals to be in the queue at once. This
parallelized version also performed fitness calculation in parallel using two threads of execution in C++.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
the NIH (R01GM115564) to KRT.
